### 2023-08-21 sections/2-methods.tex
repeated rnn mostly

In this manuscript, we are primarily concerned with $k$-nearest neighbors search in a finite-dimensional 
vector space. Some of the algorithms we present use a modified version of the $\rho$-nearest neighbors search 
algorithm described in CHESS~\cite{ishaq2019clustered} to conduct $k$-NN search. (TODO: make second section less 
vague once we determine exactly what we're doing with sharding and indexing.)

@@

Some of our algorithms for $\rho$- and $k$-nearest neighbors search rely on the \emph{local fractal dimension} at some region of the dataset, 
which we define as: 
$$\text{log}_2 (\frac{|B_D(q, r_1)|}{|B_D(q, r_2)|})$$
where $B_D(q, r)$ is the set of points contained in a ball of radius $r$ 
centered on a point $q$ in the dataset $D$; in this example, we compute fractal dimension for some radius $r_1$ and a smaller radius $r_2 = \frac{r_1}{2}$.
We stress that this concept of local fractal dimension differs from the \emph{embedding dimension} of a dataset. To illustrate the difference,
consider SDSS's APOGEE dataset, wherein each datum is a nonnegative real-valued vector of length 8,575. Hence, the \emph{embedding dimension} of this dataset is 8,575. 
However, due to physical constraints (namely, the laws that govern stellar fusion and spectral emission lines), the data are constrained to a lower-dimensional 
manifold. The \emph{local fractal dimension} is an approximation of the dimensionality of that lower-dimensional manifold at a given point, for some length scale.
The notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold is known as the \emph{manifold hypothesis}.

@@

@@

@@

@@

The $k$-nearest neighbors search problem is posed as follows: given a query $q$ and a positive integer $k$, find the $k$ closest instances in $X$.
We present five variants of $k$-nearest neighbors search: Greedy Search, K-NN by Repeated $\rho$-NN, Sieve Search I, Sieve Search II, and 
Linear Search. We use a preliminary auto-tuning function to predict the variant which will perform 
best for a given query, dataset, and value of $k$. We then proceed with search using that variant.

Given a Cluster $C$, let $c$ be its center and $r$ be its radius. Our $k$-NN variants make use of the following cluster 
properties:
\begin{itemize}
    \item $\delta = f(q, c)$ is the distance from the query to the cluster center $c$.
    \item $\delta_{max} = \delta + r$ is the distance from the query to the potentially farthest instance in $C$.
    \item $\delta_{min} = \text{max}(0, \delta - r)$ is the distance from the query to the potentially closest instance in $C$.
\end{itemize}

We sometimes also use the notation $\delta = f(x, q)$ with points instead of clusters to refer to the distance from a point $x$ to the query $q$.

@@

\begin{algorithm} 
\caption{Greedy Search} 
\label{alg:greedy_search} 
\begin{algorithmic}[1]
    \REQUIRE $tree$, $query$, $k$
    \STATE $Q \leftarrow$ priority queue
    \STATE $Q.push(tree.root)$
    \STATE $P \leftarrow$ priority queue
    \WHILE{$|P| < k \lor P.max.\delta > Q.min.\delta_{min}$}
        \WHILE{$!Q.min.isLeaf$}
            \STATE $[l, r] \leftarrow Q.extractMin().children()$
            \STATE $Q.push([l, r])$
        \ENDWHILE
        \STATE $leaf \leftarrow Q.extractMin()$
        \STATE $P.push(leaf)$
        \WHILE{$|P| > k$}
            \STATE $P.extractMax()$
        \ENDWHILE
    \ENDWHILE
    \STATE Return $P$
\end{algorithmic}
\end{algorithm}

\subsubsection{$k$-NN by Repeated $\rho$-NN}
\label{subsubsec:methods:knn-search:repeated-rnn}

Hence its name, $k$-NN by Repeated $\rho$-NN repeatedly performs 
$\rho$-nearest neighbors search, increasing the radius until $k$ neighbors
have been found.

Search starts with a radius $m$ equal to the radius of the cluster tree divided by
the cardinality of the dataset. We perform $\rho$-NN search with radius $m$. 
If no points are within a distance $m$ of the query, we increase the radius by a factor of 
2 and perform $rho$-NN search again, repeating until at least one point is found.

Once we have found at least one point, we continue to perform $\rho$-NN search, but instead of 
increasing the radius by a factor of 2 on each iteration, we increase it by a factor determined 
by the local fractal dimension of the clusters containing the points found so far. In particular, 
we increase the radius by a factor of $\text{max}(2, \frac{k}{|P|^{\frac{1}{\mu}}}$), where $|P|$ is the number of 
points found so far and $\mu$ is the mean local fractal dimension of the clusters containing the points found so far.
Since the local fractal dimension is memoized at clustering time, this poses no additional cost at search time.
Once $|P| >= k$, we sort the points in $P$ by $\delta$ and return the $k$ closest points.

TODO: 
- add intutiion for factor 
- introduce P variable sooner

\subsubsection{Sieve Search I}
\label{subsubsec:methods:knn-search:sieve}

\subsubsection{Sieve Search II}

In this variant of KNN, we repeatedly calculate a threshold distance $\tau$ such that no cluster with a $\delta_{min}$ greater than $\tau$ 
can contain one of the $k$ nearest neighbors. Search terminates when we have a threshold containing exactly $k$ points.

Let $Q$ be a vector of clusters sorted by non-decreasing $\delta_{max}$. Initialize $Q$ with  
the root cluster. 
Let $P$ be a priority queue of size $k$ containing points sorted by non-decreasing distance from the query. This queue starts empty.

Repeat the following: 
\begin{enumerate}
\item Find the cluster $C_{\tau}$ in $Q$ with the smallest $\delta_{max}$ such that no cluster whose $\delta_{min}$ is greater than $C_{\tau}$'s $\delta_{max}$ can contain one of the $k$ nearest neighbors.
\item Let the threshold $\tau$ be the $\delta_{max}$ of $C_{\tau}$.
\item Remove from $P$ any points which are a distance greater than $\tau$ from the query. 
\item If no cluster in $Q$ has a $\delta_{max}$ less than or equal to $\tau$, break.
\item Remove from $Q$ any cluster whose $\delta_{min}$ is greater than $\tau$.
\item Transfer from $Q$ to $P$ all the points in any cluster whose cardinality is less than $k$ or who is a leaf. 
\item Replace all non-leaf clusters in $Q$ with their children. 
\end{enumerate}

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Sieve V1} % give the algorithm a caption
    \label{alg:sieve_v1} % and a label for \ref{} commands later in the document
    \begin{algorithmic}[2] % enter the algorithmic environment
        \REQUIRE $tree$, $query$, $k$
        \STATE $Q \leftarrow$ priority queue
        \STATE $Q.push(tree.root)$
        \STATE $P \leftarrow$ priority queue
        \WHILE{$|P| < k \lor P.max.\delta > Q.max.\delta_{min}$}
            \WHILE{$!Q.extractMax().isLeaf$}
                \STATE $[l, r] \leftarrow Q.extractMax().children()$
                \STATE $Q.push([l, r])$
            \ENDWHILE
            \STATE $leaf \leftarrow Q.extractMax()$
            \STATE $P.push(leaf)$
            \WHILE{$|P| > k$}
                \STATE $P.extractMax()$
            \ENDWHILE
        \ENDWHILE
        \STATE Return $P$
    \end{algorithmic}
    \end{algorithm}

@@

\label{subsubsec:methods:clustering:proof-of-eventually-decreasing-cluster-radii}

\subsubsection {Complexity}

Modified binary search to identify candidate leaf clusters.
Exhaustive leaf search over those leaves.

Lift this from CHESS paper. Be sure to cite CHESS here.

Complexity is the same as in CHESS.

The k-nearest neighbors search problem is posed as follows: given a query $q$ and a positive integer $k$, find the $k$ closest instances in $X$.

Given a Cluster $C$, let $c$ be its center and $r$ be its radius. Our $k$-nn variants make use of the following definitions:
\begin{itemize}
    \item $\delta = f(q, c)$ is the distance from the query to the cluster center.
    \item $\delta_{max} = \delta + r$ is the distance from the query to the potentially farthest instance in the Cluster.
    \item $\delta_{min} = \text{max}(0, \delta - r)$ is the distance from the query to the potentially closest instance in the Cluster.
\end{itemize}

\subsubsection{Expanding Threshold}
\label{subsubsec:methods:knn-search:expanding-threshold}
In this variant of KNN, we repeatedly

Let $Q$ be a priority queue of clusters sorted by non-increasing $\delta_{min}$. Initialize $Q$ with the root cluster.
Let $P$ be a second priority queue of points sorted by non-decreasing distance from the query. This queue starts empty.

\begin{algorithm} % enter the algorithm environment
\caption{Expanding Threshold} % give the algorithm a caption
\label{alg:expanding_threshold} % and a label for \ref{} commands later in the document
\begin{algorithmic}[1] % enter the algorithmic environment
    \REQUIRE $tree$, $query$, $k$
    \STATE $Q \leftarrow$ priority queue
    \STATE $Q.push(tree.root)$
    \STATE $P \leftarrow$ priority queue
    \WHILE{$|P| < k \lor P.max.\delta > Q.min.\delta_{min}$}
        \WHILE{$!Q.min.isLeaf$}
            \STATE $[l, r] \leftarrow Q.extractMin().children()$
            \STATE $Q.push([l, r])$
        \ENDWHILE
        \STATE $leaf \leftarrow Q.extractMin()$
        \STATE $P.push(leaf)$
        \WHILE{$|P| > k$}
            \STATE $P.extractMax()$
        \ENDWHILE
    \ENDWHILE
    \STATE Return $P$
\end{algorithmic}
\end{algorithm}

\subsubsection{Sieve V1}
\label{subsubsec:methods:knn-search:sieve-v1}

In this variant of KNN, we repeatedly calculate a threshold distance $\tau$ such that no cluster with a $\delta_{min}$ greater than $\tau$ 
can contain one of the $k$ nearest neighbors. Search terminates when we have a threshold containing exactly $k$ points.

Let $Q$ be a vector of clusters sorted by non-decreasing $\delta_{max}$. Initialize $Q$ with  
the root cluster. 
Let $P$ be a priority queue of size $k$ containing points sorted by non-decreasing distance from the query. This queue starts empty.

Repeat the following: 
\begin{enumerate}
\item Find the cluster $C_{\tau}$ in $Q$ with the smallest $\delta_{max}$ such that no cluster whose $\delta_{min}$ is greater than $C_{\tau}$'s $\delta_{max}$ can contain one of the $k$ nearest neighbors.
\item Let the threshold $\tau$ be the $\delta_{max}$ of $C_{\tau}$.
\item Remove from $P$ any points which are a distance greater than $\tau$ from the query. 
\item If no cluster in $Q$ has a $\delta_{max}$ less than or equal to $\tau$, break.
\item Remove from $Q$ any cluster whose $\delta_{min}$ is greater than $\tau$.
\item Transfer from $Q$ to $P$ all the points in any cluster whose cardinality is less than $k$ or who is a leaf. 
\item Replace all non-leaf clusters in $Q$ with their children. 
\end{enumerate}

\begin{algorithm} % enter the algorithm environment
    \caption{Sieve V1} % give the algorithm a caption
    \label{alg:sieve_v1} % and a label for \ref{} commands later in the document
    \begin{algorithmic}[2] % enter the algorithmic environment
        \REQUIRE $tree$, $query$, $k$
        \STATE $Q \leftarrow$ priority queue
        \STATE $Q.push(tree.root)$
        \STATE $P \leftarrow$ priority queue
        \WHILE{$|P| < k \lor P.max.\delta > Q.max.\delta_{min}$}
            \WHILE{$!Q.extractMax().isLeaf$}
                \STATE $[l, r] \leftarrow Q.extractMax().children()$
                \STATE $Q.push([l, r])$
            \ENDWHILE
            \STATE $leaf \leftarrow Q.extractMax()$
            \STATE $P.push(leaf)$
            \WHILE{$|P| > k$}
                \STATE $P.extractMax()$
            \ENDWHILE
        \ENDWHILE
        \STATE Return $P$
    \end{algorithmic}
    \end{algorithm}

\subsubsection{Sieve V2}
\label{subsubsec:methods:knn-search:sieve-v2}
Add explanation of the sieve v2 algorithm.

\subsubsection{Repeated Rnn}
\label{subsubsec:methods:knn-search:repeated-rnn}
Add explanation of the repeated rnn algorithm.

>>>>

In this manuscript, we are primarily concerned with $k$-nearest neighbors search in a finite-dimensional 
vector space. Some of the algorithms we present use a modified version of the $\rho$-nearest neighbors search 
algorithm described in CHESS~\cite{ishaq2019clustered} to conduct $k$-NN search. (TODO: make second sentence less 
vague once we determine exactly what we're doing with sharding and indexing.)

@@

Some of our algorithms for $\rho$- and $k$-nearest neighbors search rely on the \emph{local fractal dimension} of some region of the dataset, 
which we define as: 
$$\text{log}_2 \left( \frac{|B_D(q, r_1)|}{|B_D(q, r_2)|} \right)$$
where $B_D(q, r)$ is the set of points contained in a ball of radius $r$ 
centered on a point $q$ in the dataset $D$; in this example, we compute fractal dimension for some radius $r_1$ and a smaller radius $r_2 = \frac{r_1}{2}$.
We stress that this concept of local fractal dimension differs from the \emph{embedding dimension} of a dataset. To illustrate the difference,
consider SDSS's APOGEE dataset, wherein each datum is a nonnegative real-valued vector of length 8,575. Hence, the \emph{embedding dimension} of this dataset is 8,575. 
However, due to physical constraints (namely, the laws that govern stellar fusion and spectral emission lines), the data are constrained to a lower-dimensional 
manifold. The \emph{local fractal dimension} is an approximation of the dimensionality of that lower-dimensional manifold at a given point, for some length scale.
The notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold is known as the \emph{manifold hypothesis}.

@@

Given a Cluster $C$, let $c$ be its center and $r$ be its radius. Our $\rho$- and $k$-NN algorithms make use of the following cluster 
properties:
\begin{itemize}
    \item $\delta = f(q, c)$ is the distance from the query to the cluster center $c$.
    \item $\delta_{max} = \delta + r$ is the distance from the query to the theoretically farthest possible instance in $C$.
    \item $\delta_{min} = \text{max}(0, \delta - r)$ is the distance from the query to the theoretically closest possible instance in $C$.
\end{itemize}

We sometimes also use the above notation with points instead of clusters. For example, we sometimes use $\delta = f(x, q)$ to refer to the distance 
from a point $x$ to the query $q$. Since we can think of a point as having radius 0, for points we have that $\delta_{max} = \delta = \delta_{min}$.

@@

TODO: This needs some doctoring.

@@

\begin{algorithm} 
    \caption{$\rho$-NN(\emph{clusters, query, r})} 
    \label{alg:rnn} 
    \begin{algorithmic}[2]
        \REQUIRE $r \geq 0$
        \REQUIRE $clusters \neq \emptyset$
        \STATE $results \leftarrow \emptyset$
        \IF{$clusters.left$}
            \IF{$clusters.left.\delta$ $\leq$ $r$ + $clusters.left.radius$}
                \STATE $\rho$-NN($clusters.left, query, r$)
            \ENDIF
        \ENDIF
        \IF{$clusters.right$}
            \IF{$clusters.right.\delta$ $\leq$ $r$ + $clusters.right.radius$} 
                \STATE $\rho$-NN($clusters.right, query, r$)
            \ENDIF
        \ENDIF
        \IF{$\neg$$clusters.left$ $\land \neg$$clusters.right$}
            \FOR{$p \in clusters.points$}
                \IF{$p.\delta \leq r$}
                    \STATE $results \leftarrow r$
                \ENDIF
            \ENDFOR
        \ENDIF
        \STATE Return $results$
    \end{algorithmic}
    \end{algorithm}

@@

In this section, we present four novel variants of $k$-nearest neighbors search: Greedy Search, K-NN by Repeated $\rho$-NN, Sieve Search I, and Sieve Search II. 
We use a preliminary auto-tuning function (TODO: do we though?) to predict the variant which will perform 
best for a given query, dataset, and value of $k$. We then proceed with search using that variant.

@@

\begin{algorithm} 
\caption{GreedySearch(\emph{tree, query, k})} 
\label{alg:greedy_search} 
\begin{algorithmic}[3]
    \STATE $Q \leftarrow$ priority queue
    \STATE $Q.push(tree.root)$
    \STATE $P \leftarrow$ priority queue
    \WHILE{$|P| < k \lor P.max.\delta > Q.min.\delta_{min}$}
        \WHILE{$!Q.min.isLeaf$}
            \STATE $[l, r] \leftarrow Q.extractMin().children()$
            \STATE $Q.push([l, r])$
        \ENDWHILE
        \STATE $leaf \leftarrow Q.extractMin()$
        \STATE $P.push(leaf)$
        \WHILE{$|P| > k$}
            \STATE $P.extractMax()$
        \ENDWHILE
    \ENDWHILE
    \STATE Return $P$
\end{algorithmic}
\end{algorithm}

\subsubsection{$k$-NN by Repeated $\rho$-NN}
\label{subsubsec:methods:knn-search:repeated-rnn}
Hence the name, with $k$-NN by Repeated $\rho$-NN we repeatedly perform
$\rho$-nearest neighbors search, increasing the search radius until $k$ neighbors
have been found.

Let $P$ be the set of nearest neighbors found thus far.
Search starts with a radius $m$ equal to the radius of the cluster tree divided by
the cardinality of the dataset. We perform $\rho$-NN search with radius $m$. 
If no points are within a distance $m$ of the query, we increase the radius by a factor of 
2 and perform $\rho$-NN search again, repeating until at least one point is found, i.e., 
until $|P| > 0$.

Once $|P| > 0$, we continue to perform $\rho$-NN search, but instead of 
increasing the radius by a factor of 2 on each iteration, we increase it by a factor determined 
by the local fractal dimension of the clusters containing the points found so far. In particular, 
we increase the radius by a factor of 
$$\text{min}\left(2, \frac{k}{|P|^{\frac{1}{\mu}}}\right)$$
where $\mu$ is the mean local fractal dimension (abbreviated lfd in algorithm below) of the clusters containing the points found so far.
Intuitively, the factor by which we increase the radius should be \emph{inversely} proportional to the number of points found so far. 
Additionally, when the local fractal dimension near the query and found neighbors is low, this suggests that the data 
are relatively concentrated in that region; thus, the factor of radius increase should be \emph{directly} proportional to the 
local fractal dimension. Since the local fractal dimension is memoized at clustering time, this poses no additional cost at search time.
Once $|P| >= k$, we sort the points in $P$ by $\delta$ and return the $k$ closest points.

\begin{algorithm} % enter the algorithm environment
    \caption{Repeated$\rho$-NN(\emph{tree, query, k})} % give the algorithm a caption
    \label{alg:knn-by-rnn} % and a label for \ref{} commands later in the document
    \begin{algorithmic}[4] % enter the algorithmic environment
        \STATE $P \leftarrow$ $\emptyset$
        \STATE $m \leftarrow$ $\frac{tree.radius}{tree.cardinality}$
        \WHILE {$|P| = 0$}
            \STATE $P.push(\rho$-NN$(tree, query, m)$)
            \STATE $m \leftarrow 2m$
        \ENDWHILE
        \WHILE {$|P| < k$}
            \STATE $clusters \leftarrow \{ C: \exists p \in P \land p \in C \}$
            \STATE $\mu \leftarrow \frac{1}{|clusters|} \sum_{C \in clusters} C.lfd$
            \STATE $m \leftarrow \text{min}\left(2, \frac{k}{|P|^{\frac{1}{\mu}}}\right)$
            \STATE $P.push(\rho$-NN$(tree, query, m)$)
        \ENDWHILE
        \STATE $P.sort()$
        \STATE Return $P[0..k]$
    \end{algorithmic}
    \end{algorithm}

\subsubsection{Sieve Search I}
\label{subsubsec:methods:knn-search:sieve}
With Sieve Search I, we repeatedly calculate a threshold distance $\tau$ such that no cluster with a $\delta_{min}$ greater than $\tau$ 
can contain one of the $k$ nearest neighbors. Search terminates when we have a threshold containing exactly $k$ points.

We start by letting $Q$ be a vector of clusters and points sorted by non-decreasing $\delta_{max}$ and initializing $Q$ with the root cluster. 
While we haven't found a threshold which guarantees exactly $k$ points, we do the following:

TODO: Fix this to reflect new algorithm.

\begin{enumerate}
\item Find the cluster $C_{\tau}$ in $Q$ with the smallest $\delta_{max}$ such that no cluster whose $\delta_{min}$ is greater than $C_{\tau}$'s $\delta_{max}$ can contain one of the $k$ nearest neighbors.
\item Let the threshold $\tau$ be the $\delta_{max}$ of $C_{\tau}$.
\item Remove from $Q$ any cluster whose $\delta_{min}$ is greater than $\tau$.
\item Transfer from $Q$ to $P$ all the points in any cluster whose cardinality is less than $k$ or who is a leaf. 
\item Replace all non-leaf clusters in $Q$ with their children. 
\end{enumerate}

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Sieve } % give the algorithm a caption
    \label{alg:sieve_v1} % and a label for \ref{} commands later in the document
    \begin{algorithmic}[2] % enter the algorithmic environment
        \REQUIRE $tree$, $query$, $k$
        \STATE $Q \leftarrow$ priority queue
        \STATE $Q.push(tree.root)$
        \STATE $P \leftarrow$ priority queue
        \WHILE{$|P| < k \lor P.max.\delta > Q.max.\delta_{min}$}
            \WHILE{$!Q.extractMax().isLeaf$}
                \STATE $[l, r] \leftarrow Q.extractMax().children()$
                \STATE $Q.push([l, r])$
            \ENDWHILE
            \STATE $leaf \leftarrow Q.extractMax()$
            \STATE $P.push(leaf)$
            \WHILE{$|P| > k$}
                \STATE $P.extractMax()$
            \ENDWHILE
        \ENDWHILE
        \STATE Return $P$
    \end{algorithmic}
    \end{algorithm}

\subsubsection{Sieve Search II}
\label{subsubsec:methods:knn-search:sieve2}

@@
